MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-28 main/MSstatsPTM_Main.tex
moved summation anova section

The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
\begin{eqnarray}
\hat{u}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}

Where $\hat{u}^*_{ij}$ is the run-level summarization of modified features. For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, $\hat{u}^*_{ij}$ is

\begin{eqnarray}
\hat{u}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
\label{eq:anova_summ}
\end{eqnarray}

@@

\medskip \noindent 
The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, and considering sample-wise ratios of the summaries. On the log$_2$ scale, for condition $i$ and biological replicate $j$ the approach is equivalent to calculating
\begin{eqnarray}
\hat{u}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values with the basic ANOVA, which corresponds to 
\begin{eqnarray}
\hat{u}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.

>>>>

First run-level summarization of modified features is performed. This is generally done by taking the log sum of the original modified features. For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, the summarization would be

\begin{eqnarray}
\hat{y}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
\label{eq:anova_summ}
\end{eqnarray}

Then the basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
\begin{eqnarray}
\hat{u}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}

@@

\medskip \noindent 
The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, and considering sample-wise ratios of the summaries. On the log$_2$ scale, for condition $i$ and biological replicate $j$ the approach is equivalent to calculating
\begin{eqnarray}
\hat{y}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values with the basic ANOVA, which corresponds to 
\begin{eqnarray}
\hat{y}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
